Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-04-12 manuscript.qmd
start writing on data section

## Incorporating Midline Crossing

@@

$$
\begin{aligned}
P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
P(\epsilon = \texttt{True} \mid t) &= 1 - (1 - p_\epsilon)^t
\end{aligned}
$$

@@

```{python}
#| echo: false
#| label: fig-midext-evo
#| fig-cap: The evolution of the probability for the tumor to remain lateralized (green) or to cross the mid-sagittal plane (red).
t = np.linspace(0, 10, 11)
p_midline = {
  "lateralized": (1 - midline_model.midext_prob)**t,
  "extension": 1 - (1 - midline_model.midext_prob)**t,
}
dist = {
  "early": midline_model.get_distribution("early").pmf,
  "late": midline_model.get_distribution("late").pmf,
}
p_colors = {"lateralized": COLORS["green"], "extension": COLORS["red"]}
t_colors = {"early": COLORS["blue"], "late": COLORS["orange"]}
markers = {}

@@

for label, p in p_midline.items():
  axes[0].plot(
    t, p_midline[label], "o-",
    label=label,
    color=p_colors[label],
  )
  for i, t_stage in enumerate(["early", "late"]):
    if label == "lateralized":
      axes[0].bar(
        t + i*w - w/2, dist[t_stage],
        color=t_colors[t_stage],
        width=w,
        label=f"{t_stage} T-cat.",
      )
    axes[1].plot(
      t, p * dist[t_stage], "o-",
      c=p_colors[label],
      mfc=t_colors[t_stage],
      mec=t_colors[t_stage],
      label=f"{label}, {t_stage}",
    )

@@

axes[1].set_ylim(0., 0.25)
axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[1].set_ylabel("probability")
axes[1].grid(axis="y", color=COLORS["gray"], alpha=0.5)

@@

1. We assume the spread *among* the LNLs to be same on both sides. It is reasonable to assume the lymphatic system is symmetric. Thus, the spread rates from one LNL to the other should be symmetric, too. Formally, this means \
   $$
   \begin{aligned}
   b_v^\text{c} &\neq b_v^\text{i} \\
   t_{rv}^\text{c} &= t_{rv}^\text{i}
   \end{aligned}
   $$ {#eq-symmetries}
   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
2. The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side. This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to its mirror position. In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed. If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes. We encode this in a *mixing parameter* $\alpha$ that captures a "degree of asymmetry": \
   $$
   b_v^\text{c,e} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\centernot{\text{e}}}
   $$ {#eq-mixing}

## Tumors Extending Over the Mid-Sagittal Line {#sec-midline}

- higher prevalence of contralateral involvement when tumor crosses the mid-sagittal plane
- but still lower than on ipsilateral side
- spread to contralateral side in case of midline extension may be describes as linear combination of ipsi- and contralateral spread in case of no midline extension.
- only one additional "mixing parameter" $\alpha$
- tumor's midline extension modelled as a binary RV (lateralized or crossing/extending)

@@

- describe origin of data
- why is this data special?
- mention [LyProX](https://lyprox.org)
- visualize data, e.g. using upsetplots

>>>>

## Modelling Midline Extension {#sec-midline}

@@

$$
\begin{aligned}
P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
P(\epsilon = \texttt{True} \mid t) &= 1 - P(\epsilon = \texttt{False} \mid t)
\end{aligned}
$$

@@

```{python}
#| echo: false
#| label: fig-midext-evo
#| fig-cap: The top panel shows the prior probability to get diagnosed at time-step $t$ for early and late T-category tumors as bars. Also in the top panel, we plot the conditional probability of the tumor's midline extension ($\epsilon=\texttt{True}$), given the time-step $t$ as a line plot. In the bottom panel, we show the joint probability of getting diagnosed in time-step $t$ *and* having a tumor that crosses the midline.
t = np.linspace(0, 10, 11)
p_midline = {
  "lateralized": (1 - midline_model.midext_prob)**t,
  "extension": 1 - (1 - midline_model.midext_prob)**t,
}
dist = {
  "early": midline_model.get_distribution("early").pmf,
  "late": midline_model.get_distribution("late").pmf,
}
p_colors = {"lateralized": COLORS["green"], "extension": COLORS["red"]}
t_colors = {"early": COLORS["blue"], "late": COLORS["orange"]}
markers = {}

@@

for label, p in p_midline.items():
  axes[0].plot(
    t, p_midline[label], "o-",
    label=f"cond. prob. $P(\epsilon={label=='extension'} \mid t)$",
    color=p_colors[label],
  )
  for i, t_stage in enumerate(["early", "late"]):
    if label == "lateralized":
      axes[0].bar(
        t + i*w - w/2, dist[t_stage],
        color=t_colors[t_stage],
        width=w,
        label=f"{t_stage} T-cat. prior $P(t)$",
      )
    axes[1].plot(
      t, p * dist[t_stage], "o-",
      c=p_colors[label],
      mfc=t_colors[t_stage],
      mec=t_colors[t_stage],
      label=f"{label} ($\epsilon={label=='extension'}$) for {t_stage} T-cat.",
    )

@@

axes[1].set_ylim(0., 0.25)
axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[1].set_ylabel("joint probability $P(\epsilon, t)$")
axes[1].grid(axis="y", color=COLORS["gray"], alpha=0.5)

@@

1. We assume the spread *among* the LNLs to be same on both sides. It is reasonable to assume the lymphatic system is symmetric. Thus, the spread rates from one LNL to the other should be symmetric, too. Formally, this means \
   $$
   \begin{aligned}
   b_v^\text{c} &\neq b_v^\text{i} \\
   t_{rv}^\text{c} &= t_{rv}^\text{i}
   \end{aligned}
   $$ {#eq-symmetries}
   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
2. The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side. This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side. In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed. If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes. We encode this in a *mixing parameter* $\alpha$ that captures a "degree of asymmetry": \
   $$
   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
   $$ {#eq-mixing}
   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline. Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.

Together with the explicit modelling of the tumor's midline extension $\epsilon$ from @sec-midline, we now have a model that may be capable of capturing the higher prevalence of contralateral involvment that comes with tumors extending over the mid-sagittal line.

@@

The data used to train our models consists of <!-- `{python} len(data)` --> patients with newly diagnosed oropharyngeal squamous cell carcinomas. It reports the involvement of every patient individually and per lymph node level in tabular form, in addition to other clinico-pathological information such as age, T-category, and HPV p16 status. Their patient records have been collected at four different instituions:

1. <!-- `{python} len(usz_data)` --> patients from the University Hospital Zurich [@ludwig_dataset_2022].
2.

- why is this data special?
- mention [LyProX](https://lyprox.org)
- visualize data, e.g. using upsetplots
